---
title: "Identifying Harmful and Costly Weather Events"
output: 
  html_document:
    keep_md: yes

---
# Synopsis
In this analysis, I aim to describe the most dangerous weather events to population health, as described by fatalities and injuries, and the economoby. My hypothesis is that tornadoes are the most dangerous to population health and the economy as they are difficult to predict and can cause significant damage. To investigate this hypothesis, data published by the National Oceanic and Atmospheric Administration documenting significant weather events including:
1) Storms with enough intensity to affect population health or cause property damage,
2) Weather anomalies such as snow flurries in Miami
3) Record maximum and minimum temperatures or precipitation occuring in connection with other weather events. 

From these data, I found tornadoes to be significantly more detrimental to population health as opposed to other weather events. 

# Loading and Pre-processing of Data
Data on weather events was downloaded from https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2. Weather events were classified based on the documentation provided here: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf

### Loading necessary libraries
dplyr, fuzzyjoin, and ggplot 2 libraries are loaded. Additionally, the knitr options are established.
```{r global_options, include=TRUE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='figures/',
                      echo=TRUE, warning=FALSE, message=FALSE)
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(fuzzyjoin))
suppressPackageStartupMessages(library(ggplot2))
```

### Download and load file
The file is downloaded if not already present in the working directory and loaded into the variable stormdata. The first 5 observations are printed.
```{r}
destfile<-"stormdata.csv.bz2"
fileURL<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if(!file.exists(destfile)){
        download.file(fileURL,destfile,method=auto)
}

stormdata<-read.csv("stormdata.csv.bz2")
head(stormdata,5)
```

### Create event categories
Event categories created based on categories provided in National Weather Service Instruction 10-605 (August 17, 2007). 

```{r}
eventcat<-c("Astronomical Low Tide","Avalanche","Blizzard","Coastal Flood","Cold/Wind Chill","Debris Flow","Dense Fog","Dense Smoke","Drought","Dust Devil","Dust Storm","Excessive Heat","Extreme Cold/Wind Chill","Flash Flood","Flood","Frost/Freeze","Funnel Cloud","Freezing Fog","Hail","Heat","Heavy Rain","Heavy Snow","High Surf","High Wind","Hurricane (Typhoon)","Ice Storm","Lake-Effect Snow","Lakeshore Flood","Lightning","Marine Hail","Marine High Wind","Marine Strong Wind","Marine Thunderstorm Wind","Rip Current","Seiche","Sleet","Storm Surge/Tide","Strong Wind","Thunderstorm Wind","Tornado","Tropical Depression","Tropical Storm","Tsunami","Volcanic Ash","Waterspout","Wildfire","Winter Storm","Winter Weather")
eventcat<-as_tibble(eventcat)
names(eventcat)<-"EVTYPE"
```


### Subset necessary columns
Subset EVTYPE, FATALITIES, and INJURIES to help determine overall mortality of weather events. The event categories are fuzzyjoined to this subset. A marker column is added to keep track if a category has been assigned to the event type. Then the subset is summarized by event category and event type with the health indicator added as the sum of the fatalities and injuries. This table is arranged in descending order by the health column.
```{r}
eventtype<-stormdata %>% select(EVTYPE,FATALITIES,INJURIES) %>% transform(EVTYPE=tolower(EVTYPE)) %>% arrange(EVTYPE)
eventtype.table<-as_tibble(eventtype)
eventtype<-eventtype.table %>% stringdist_left_join(eventcat,method="lcs",ignore_case=TRUE,max_dist=1)

names(eventtype)<-c("EVTYPE","FATALITIES","INJURIES","EVCAT")

eventtype[,"n"]<-1
eventtype$n[is.na(eventtype$EVCAT)]<-0

#Summarize
eventtype<-eventtype %>% group_by(EVCAT,EVTYPE) %>% summarize(FATALITIES=sum(FATALITIES),INJURIES=sum(INJURIES),health=sum(FATALITIES,INJURIES,na.rm=TRUE),n=sum(n)) %>% arrange(desc(health))

#Remove unnecessary data frames
rm(stormdata, eventtype.table,eventcat)

```

### Clean the event types to correspond to the 48 listed types
Collapse the event types into appropriate groups. To do this, Add an event category column to avoid changing the raw data. Use regular expression functions to identify the categories, and set the category column for those events to one of the 48 categories listed in the accompanying documentation. Any event that could not be easily classified was defined as 'other'. To classify each category, the category pattern was selected and viewed. If one occurrence had multiple events, it was classifed by the most likely cause of injury or destruction. For example, excessive heat causes more injuries and fatalities than drought, and therefore excessive heat/drought was classified as excessive heat. Tornadoes are more destructive than thunderstorm wind and hail, so 'tornadoes, tstm wind, hail' was classified as tornado. Categories were classified by subsetting the unique eventtypes with the category. The EVCAT was set equal to the appropriate category and n was changed to 1.The subset was then removed. 

* **Notes:**
+ Landslide, mudslide, and rock slide combined with avalanche
+ Blow-out tide added to astronomical low tide
+ Excessive heat/drought classified as excessive heat
+ Cold/Wind Chill combined with Extreme Cold/Wind Chill
+ Heavy snow/ice storm classified as heavy snow
+ The wind categories should be run last to avoid conflicts
+ All winter events not included in previous categories were grouped into Winter Weather
+ Cold, Dry, High Seas, Wind were added as categories


```{r}
#adjust avalanche and landslide
avalanche<-unique(eventtype$EVTYPE[grepl("avalan",eventtype$EVTYPE, ignore.case=TRUE)|grepl("*slide*",eventtype$EVTYPE, ignore.case=TRUE)])
eventtype$EVCAT[which(eventtype$EVTYPE %in% avalanche & eventtype$n==0)]<-"Avalanche"
eventtype$n[which(eventtype$EVTYPE %in% avalanche & eventtype$n==0)]<-1
rm(avalanche)

#adjust astronomical low tide, add blow-out tide
low.tide<-unique(eventtype$EVTYPE[grepl("low tide",eventtype$EVTYPE, ignore.case=TRUE)|grepl("blow-out",eventtype$EVTYPE, ignore.case=TRUE)])
eventtype$EVCAT[which(eventtype$EVTYPE %in% low.tide & eventtype$n==0)]<-"Astronomical Low Tide"
eventtype$n[which(eventtype$EVTYPE %in% low.tide & eventtype$n==0)]<-1
rm(low.tide)

#adjust blizzard
blizzard<-unique(eventtype$EVTYPE[grepl("*blizzard*",eventtype$EVTYPE, ignore.case=TRUE)])
eventtype$EVCAT[which(eventtype$EVTYPE %in% blizzard & eventtype$n==0)]<-"Blizzard"
eventtype$n[which(eventtype$EVTYPE %in% blizzard & eventtype$n==0)]<-1
rm(blizzard)

#adjust coastal flooding
coastal.flood<-unique(eventtype$EVTYPE[grepl("coastal",eventtype$EVTYPE, ignore.case=TRUE)|grepl("beach",eventtype$EVTYPE, ignore.case=TRUE)])
coastal.storm<-coastal.flood[grepl("storm",coastal.flood)]
coastal.flood<-coastal.flood[!grepl("storm",coastal.flood)]
eventtype$EVCAT[which(eventtype$EVTYPE %in% coastal.flood & eventtype$n==0)]<-"Coastal Flood"
eventtype$EVCAT[which(eventtype$EVTYPE %in% coastal.storm & eventtype$n==0)]<-"Coastal Storm"
eventtype$n[which(eventtype$EVTYPE %in% coastal.flood & eventtype$n==0)]<-1
eventtype$n[which(eventtype$EVTYPE %in% coastal.storm & eventtype$n==0)]<-1
rm(coastal.flood,coastal.storm)

#cold/wind chill
wind.chill<-unique(eventtype$EVTYPE[grepl("wind chill",eventtype$EVTYPE, ignore.case=TRUE) | grepl("extreme cold",eventtype$EVTYPE, ignore.case=TRUE) | grepl("extremewind",eventtype$EVTYPE, ignore.case=TRUE) | grepl("windchill",eventtype$EVTYPE, ignore.case=TRUE) | grepl("record low",eventtype$EVTYPE)])
wind.chill<-wind.chill[!grepl("blizzard",wind.chill) & !grepl("tornado",wind.chill) & !grepl("fog",wind.chill) & !grepl("frost",wind.chill)]
eventtype$EVCAT[which(eventtype$EVTYPE %in% wind.chill & eventtype$n==0)]<-"Extreme Cold/Wind Chill"
eventtype$n[which(eventtype$EVTYPE %in% wind.chill & eventtype$n==0)]<-1
rm(wind.chill)

#No debris flow

#Dense Fog, freezing fog
fog<-unique(eventtype$EVTYPE[grepl("fog",eventtype$EVTYPE, ignore.case=TRUE)])
freeze.fog<-fog[grepl("freezing",fog)|grepl("ice",fog)]
fog<-fog[!grepl("freezing",fog)&!grepl("ice",fog)]
eventtype$EVCAT[which(eventtype$EVTYPE %in% fog & eventtype$n==0)]<-"Dense Fog"
eventtype$EVCAT[which(eventtype$EVTYPE %in% freeze.fog & eventtype$n==0)]<-"Freezing Fog"
eventtype$n[which(eventtype$EVTYPE %in% fog & eventtype$n==0)]<-1
eventtype$n[which(eventtype$EVTYPE %in% freeze.fog & eventtype$n==0)]<-1
rm(fog,freeze.fog)

#Dense Smoke
smoke<-unique(eventtype$EVTYPE[grepl("smoke",eventtype$EVTYPE, ignore.case=TRUE)])
eventtype$EVCAT[which(eventtype$EVTYPE %in% smoke & eventtype$n==0)]<-"Dense Smoke"
eventtype$n[which(eventtype$EVTYPE %in% smoke & eventtype$n==0)]<-1
rm(smoke)

#drought, excessive heat/drought classified as excessive heat
drought<-unique(eventtype$EVTYPE[grepl("drought",eventtype$EVTYPE, ignore.case=TRUE) | grepl("abnormally dry",eventtype$EVTYPE, ignore.case=TRUE) | grepl("excessively dry",eventtype$EVTYPE, ignore.case=TRUE) | grepl("record dry",eventtype$EVTYPE, ignore.case=TRUE)])
drought<-drought[!grepl("excessive heat",drought)]
eventtype$EVCAT[which(eventtype$EVTYPE %in% drought & eventtype$n==0)]<-"Drought"
eventtype$n[which(eventtype$EVTYPE %in% drought & eventtype$n==0)]<-1
rm(drought)


#dust devil and dust storm
dust.devil<-unique(eventtype$EVTYPE[grepl("dust",eventtype$EVTYPE, ignore.case=TRUE)])
dust.storm<-dust.devil[grepl("storm",dust.devil)]
dust.devil<-dust.devil[!grepl("storm",dust.devil)]

eventtype$EVCAT[which(eventtype$EVTYPE %in% dust.devil & eventtype$n==0)]<-"Dust Devil"
eventtype$EVCAT[which(eventtype$EVTYPE %in% dust.storm & eventtype$n==0)]<-"Dust Storm"
eventtype$n[which(eventtype$EVTYPE %in% dust.devil & eventtype$n==0)]<-1
eventtype$n[which(eventtype$EVTYPE %in% dust.storm & eventtype$n==0)]<-1
rm(dust.devil,dust.storm)

#Excessive heat
#adjust excessive heat
heat<-unique(eventtype$EVTYPE[grepl("heat",eventtype$EVTYPE, ignore.case=TRUE) | grepl("abnormal warmth",eventtype$EVTYPE, ignore.case=TRUE) | grepl("record high",eventtype$EVTYPE)])
eventtype$EVCAT[which(eventtype$EVTYPE %in% heat & eventtype$n==0)]<-"Excessive Heat"
eventtype$n[which(eventtype$EVTYPE %in% heat & eventtype$n==0)]<-1
rm(heat)

#Flash flood, flood, lakeshore flood
flood<-unique(eventtype$EVTYPE[grepl("flood",eventtype$EVTYPE, ignore.case=TRUE) | grepl("fld",eventtype$EVTYPE, ignore.case=TRUE) | grepl("dam",eventtype$EVTYPE, ignore.case=TRUE) | grepl("flash",eventtype$EVTYPE)])
flash.flood<-flood[grepl("flash",flood)]
lake.flood<-flood[grepl("lake",flood)]
flood<-flood[!grepl("coastal",flood)&!grepl("flash",flood)& !grepl("red",flood)& !grepl("beach",flood)&!grepl("lake",flood)]
eventtype$EVCAT[which(eventtype$EVTYPE %in% flood & eventtype$n==0)]<-"Flood"
eventtype$EVCAT[which(eventtype$EVTYPE %in% flash.flood & eventtype$n==0)]<-"Flash Flood"
eventtype$EVCAT[which(eventtype$EVTYPE %in% lake.flood & eventtype$n==0)]<-"Lakeshore Flood"

eventtype$n[which(eventtype$EVTYPE %in% flood & eventtype$n==0)]<-1
eventtype$n[which(eventtype$EVTYPE %in% flash.flood & eventtype$n==0)]<-1
eventtype$n[which(eventtype$EVTYPE %in% lake.flood & eventtype$n==0)]<-1

rm(flood,flash.flood,lake.flood)

#frost/freeze
frost<-unique(eventtype$EVTYPE[grepl("frost",eventtype$EVTYPE, ignore.case=TRUE) | grepl("freez",eventtype$EVTYPE, ignore.case=TRUE)])
frost<-frost[!grepl("fog",frost) & !grepl("blizzard",frost) & !grepl("rain",frost) & !grepl("sleet",frost) & !grepl("record cold",frost)]
eventtype$EVCAT[which(eventtype$EVTYPE %in% frost & eventtype$n==0)]<-"Frost/Freeze"
eventtype$n[which(eventtype$EVTYPE %in% frost & eventtype$n==0)]<-1
rm(frost)

#Funnel Cloud
funnel<-unique(eventtype$EVTYPE[grepl("funnel",eventtype$EVTYPE)])
funnel<-funnel[!grepl("wind",funnel) & !grepl("water",funnel) & !grepl("cold",funnel) & !grepl("hail",funnel)]
eventtype$EVCAT[which(eventtype$EVTYPE %in% funnel & eventtype$n==0)]<-"Funnel Cloud"
eventtype$n[which(eventtype$EVTYPE %in% funnel & eventtype$n==0)]<-1
rm(funnel)

#Hail
hail<-unique(eventtype$EVTYPE[grepl("hail",eventtype$EVTYPE)])
marine.hail<-hail[grepl("marine",hail)]
hail<-hail[!grepl("flood",hail) & !grepl("marine",hail) & !grepl("tornado",hail)]
eventtype$EVCAT[which(eventtype$EVTYPE %in% hail & eventtype$n==0)]<-"Hail"
eventtype$EVCAT[which(eventtype$EVTYPE %in% marine.hail & eventtype$n==0)]<-"Marine Hail"

eventtype$n[which(eventtype$EVTYPE %in% hail & eventtype$n==0)]<-1
eventtype$n[which(eventtype$EVTYPE %in% marine.hail & eventtype$n==0)]<-1
rm(hail,marine.hail)

#heavy rain
rain<-unique(eventtype$EVTYPE[grepl("rain",eventtype$EVTYPE) | grepl("precip",eventtype$EVTYPE) | grepl("*shower*",eventtype$EVTYPE)])
rain<-rain[!grepl("flood",rain) & !grepl("blizzard",rain) & !grepl("freez",rain) & !grepl("lightning",rain) & !grepl("record low",rain) & !grepl("sleet",rain) & !grepl("thunderstorm",rain) & !grepl("tstm",rain) & !grepl("surf",rain) & !grepl("below",rain)]
eventtype$EVCAT[which(eventtype$EVTYPE %in% rain & eventtype$n==0)]<-"Heavy Rain"
eventtype$n[which(eventtype$EVTYPE %in% rain & eventtype$n==0)]<-1
rm(rain)

#heavy snow
snow<-unique(eventtype$EVTYPE[grepl("snow",eventtype$EVTYPE) | grepl("snowstorm",eventtype$EVTYPE)])
snow<-snow[grepl("heavy",snow) | grepl("excessive",snow) | grepl("record",snow) | grepl("snowstorm",snow)]
snow<-snow[!grepl("lake",snow) & !grepl("blizzard",snow)]
eventtype$EVCAT[which(eventtype$EVTYPE %in% snow & eventtype$n==0)]<-"Heavy Snow"
eventtype$n[which(eventtype$EVTYPE %in% snow & eventtype$n==0)]<-1
rm(snow)

#high tide
high.tide<-unique(eventtype$EVTYPE[grepl("high tide",eventtype$EVTYPE)])
eventtype$EVCAT[which(eventtype$EVTYPE %in% high.tide & eventtype$n==0)]<-"High Tide"
eventtype$n[which(eventtype$EVTYPE %in% high.tide & eventtype$n==0)]<-1
rm(high.tide)

#High surf
surf<-unique(eventtype$EVTYPE[grepl("surf",eventtype$EVTYPE)|grepl("swell",eventtype$EVTYPE)])
surf<-surf[!grepl("rip",surf)]
eventtype$EVCAT[which(eventtype$EVTYPE %in% surf & eventtype$n==0)]<-"High Surf"
eventtype$n[which(eventtype$EVTYPE %in% surf & eventtype$n==0)]<-1
rm(surf)

#adjust hurricane
hurricane<-unique(eventtype$EVTYPE[grepl("hurricane", eventtype$EVTYPE)])
hurricane<-hurricane[!grepl("swell",hurricane)]
eventtype$EVCAT[which(eventtype$EVTYPE %in% hurricane & eventtype$n==0)]<-"Hurricane"
eventtype$n[which(eventtype$EVTYPE %in% hurricane & eventtype$n==0)]<-1
rm(hurricane)


#ice
ice<-unique(eventtype$EVTYPE[grepl("ice storm",eventtype$EVTYPE)])
ice<-ice[!grepl("heavy snow",ice) & !grepl("flood",ice)]
eventtype$EVCAT[which(eventtype$EVTYPE %in% ice & eventtype$n==0)]<-"ice storm"
eventtype$n[which(eventtype$EVTYPE %in% ice & eventtype$n==0)]<-1
rm(ice)

#lake-effect snow
lake.snow<-unique(eventtype$EVTYPE[grepl("lake",eventtype$EVTYPE) & grepl("snow",eventtype$EVTYPE)])
eventtype$EVCAT[which(eventtype$EVTYPE %in% lake.snow & eventtype$n==0)]<-"Lake-Effect Snow"
eventtype$n[which(eventtype$EVTYPE %in% lake.snow & eventtype$n==0)]<-1
rm(lake.snow)

#lightning
lightning<-unique(eventtype$EVTYPE[grepl("light",eventtype$EVTYPE)|grepl("ning",eventtype$EVTYPE)])
lightning<-lightning[!grepl("freezing",lightning) & !grepl("snow",lightning) & !grepl("northern",lightning)]
eventtype$EVCAT[which(eventtype$EVTYPE %in% lightning & eventtype$n==0)]<-"Lightning"
eventtype$n[which(eventtype$EVTYPE %in% lightning & eventtype$n==0)]<-1
rm(lightning)

#marine
marine<-unique(eventtype$EVTYPE[grepl("marine",eventtype$EVTYPE)])
marine.other<-marine[grepl("accident",marine) | grepl("mishap",marine)]
marine.high.wind<-marine[grepl("high wind",marine)]
marine.strong.wind<-marine[grepl("strong wind",marine)]
marine.thunderstorm<-marine[grepl("thunderstorm",marine) | grepl("tstm",marine)]

eventtype$EVCAT[which(eventtype$EVTYPE %in% marine.other & eventtype$n==0)]<-"Marine Other"
eventtype$EVCAT[which(eventtype$EVTYPE %in% marine.high.wind & eventtype$n==0)]<-"Marine High Wind"
eventtype$EVCAT[which(eventtype$EVTYPE %in% marine.strong.wind & eventtype$n==0)]<-"Marine Strong Wind"
eventtype$EVCAT[which(eventtype$EVTYPE %in% marine.thunderstorm & eventtype$n==0)]<-"Marine Thunderstorm Wind"

eventtype$n[which(eventtype$EVTYPE %in% marine.other & eventtype$n==0)]<-1
eventtype$n[which(eventtype$EVTYPE %in% marine.high.wind & eventtype$n==0)]<-1
eventtype$n[which(eventtype$EVTYPE %in% marine.strong.wind & eventtype$n==0)]<-1
eventtype$n[which(eventtype$EVTYPE %in% marine.thunderstorm & eventtype$n==0)]<-1

rm(marine,marine.other,marine.high.wind,marine.strong.wind,marine.thunderstorm)


#rip current
rip.current<-unique(eventtype$EVTYPE[grepl("rip",eventtype$EVTYPE)])
eventtype$EVCAT[which(eventtype$EVTYPE %in% rip.current & eventtype$n==0)]<-"Rip Current"
eventtype$n[which(eventtype$EVTYPE %in% rip.current & eventtype$n==0)]<-1
rm(rip.current)

#seiche
seiche<-unique(eventtype$EVTYPE[grepl("seiche",eventtype$EVTYPE)])
eventtype$EVCAT[which(eventtype$EVTYPE %in% seiche & eventtype$n==0)]<-"Seiche"
eventtype$n[which(eventtype$EVTYPE %in% seiche & eventtype$n==0)]<-1
rm(seiche)

#sleet
sleet<-unique(eventtype$EVTYPE[grepl("sleet",eventtype$EVTYPE)])
sleet<-sleet[!grepl("ice storm",sleet)]
eventtype$EVCAT[which(eventtype$EVTYPE %in% sleet & eventtype$n==0)]<-"Sleet"
eventtype$n[which(eventtype$EVTYPE %in% sleet & eventtype$n==0)]<-1
rm(sleet)

#storm surge
surge<-unique(eventtype$EVTYPE[grepl("storm surge",eventtype$EVTYPE)])
eventtype$EVCAT[which(eventtype$EVTYPE %in% surge & eventtype$n==0)]<-"Storm Surge/Tide"
eventtype$n[which(eventtype$EVTYPE %in% surge & eventtype$n==0)]<-1
rm(surge)

#Thunderstorm
thunderstorm<-unique(eventtype$EVTYPE[grepl("thunderstorm",eventtype$EVTYPE)|grepl("tstm",eventtype$EVTYPE)|grepl("orm wind",eventtype$EVTYPE)|grepl("om winds",eventtype$EVTYPE)])
thunderstorm<-thunderstorm[!grepl("marine",thunderstorm) & !grepl("non",thunderstorm) & !grepl("hail",thunderstorm)]
eventtype$EVCAT[which(eventtype$EVTYPE %in% thunderstorm & eventtype$n==0)]<-"Thunderstorm Wind"
eventtype$n[which(eventtype$EVTYPE %in% thunderstorm & eventtype$n==0)]<-1
rm(thunderstorm)

#Tornado
tornado<-unique(eventtype$EVTYPE[grepl("*tornado*",eventtype$EVTYPE)])
tornado<-tornado[!grepl("water",tornado)]
eventtype$EVCAT[which(eventtype$EVTYPE %in% tornado & eventtype$n==0)]<-"Tornado"
eventtype$n[which(eventtype$EVTYPE %in% tornado & eventtype$n==0)]<-1
rm(tornado)

#tropical
tropical<-unique(eventtype$EVTYPE[grepl("tropical",eventtype$EVTYPE)])
tropical.depression<-tropical[grepl("depression",tropical)]
tropical.storm<-tropical[grepl("storm",tropical)]
eventtype$EVCAT[which(eventtype$EVTYPE %in% tropical.depression & eventtype$n==0)]<-"Tropical Depression"
eventtype$EVCAT[which(eventtype$EVTYPE %in% tropical.storm & eventtype$n==0)]<-"Tropical Storm"
eventtype$n[which(eventtype$EVTYPE %in% tropical.depression & eventtype$n==0)]<-1
eventtype$n[which(eventtype$EVTYPE %in% tropical.storm & eventtype$n==0)]<-1
rm(tropical.storm,tropical.depression,tropical)

#tsunami
tsunami<-unique(eventtype$EVTYPE[grepl("tsunami",eventtype$EVTYPE)])
eventtype$EVCAT[which(eventtype$EVTYPE %in% tsunami & eventtype$n==0)]<-"Tsunami"
eventtype$n[which(eventtype$EVTYPE %in% tsunami & eventtype$n==0)]<-1
rm(tsunami)

#volcanic eruption/ash
volcano<-unique(eventtype$EVTYPE[grepl("volcan",eventtype$EVTYPE)])
eventtype$EVCAT[which(eventtype$EVTYPE %in% volcano & eventtype$n==0)]<-"Volcanic Ash"
eventtype$n[which(eventtype$EVTYPE %in% volcano & eventtype$n==0)]<-1
rm(volcano)

#waterspout
waterspout<-unique(eventtype$EVTYPE[grepl("waterspout",eventtype$EVTYPE)])
waterspout<-waterspout[!grepl("dust",waterspout)]
eventtype$EVCAT[which(eventtype$EVTYPE %in% waterspout & eventtype$n==0)]<-"Waterspout"
eventtype$n[which(eventtype$EVTYPE %in% waterspout & eventtype$n==0)]<-1
rm(waterspout)

#wildfire
wildfire<-unique(eventtype$EVTYPE[grepl("fire",eventtype$EVTYPE)])
wildfire<-wildfire[!grepl("red",wildfire) & !grepl("lightning",wildfire)]
eventtype$EVCAT[which(eventtype$EVTYPE %in% wildfire & eventtype$n==0)]<-"Wildfire"
eventtype$n[which(eventtype$EVTYPE %in% wildfire & eventtype$n==0)]<-1
rm(wildfire)

#winter storm
winter.storm<-unique(eventtype$EVTYPE[grepl("winter storm",eventtype$EVTYPE)])
winter.storm<-winter.storm[!grepl("blizzard",winter.storm) & !grepl("heavy snow",winter.storm)]
eventtype$EVCAT[which(eventtype$EVTYPE %in% winter.storm & eventtype$n==0)]<-"Winter Storm"
eventtype$n[which(eventtype$EVTYPE %in% winter.storm & eventtype$n==0)]<-1
rm(winter.storm)

#winter weather
winter.weather<-unique(eventtype$EVTYPE[grepl("winter",eventtype$EVTYPE)| grepl("snow",eventtype$EVTYPE) | grepl("freezing",eventtype$EVTYPE) | grepl("ice",eventtype$EVTYPE)| grepl("icy",eventtype$EVTYPE) | grepl("glaze",eventtype$EVTYPE) | grepl("mix",eventtype$EVTYPE)])
winter.weather<-winter.weather[!grepl("blizzard",winter.weather) & !grepl("fog",winter.weather) & !grepl("sleet",winter.weather) & !grepl("heavy",winter.weather) & !grepl("flood",winter.weather) & !grepl("chill",winter.weather) & !grepl("storm",winter.weather) & !grepl("lake",winter.weather) & !grepl("excessive",winter.weather) & !grepl("record",winter.weather) & !grepl("storm",winter.weather) & !grepl("lack of",winter.weather)]

eventtype$EVCAT[which(eventtype$EVTYPE %in% winter.weather & eventtype$n==0)]<-"Winter Weather"
eventtype$n[which(eventtype$EVTYPE %in% winter.weather & eventtype$n==0)]<-1
rm(winter.weather)

#High wind and strong wind
wind<-unique(eventtype$EVTYPE[grepl("wind",eventtype$EVTYPE)| grepl("microburst",eventtype$EVTYPE) | grepl("downburst",eventtype$EVTYPE)])
high.wind<-wind[grepl("high wind",wind) | grepl("high  winds",wind)]
strong.wind<-wind[grepl("strong wind",wind) | grepl("microburst",wind) | grepl("downburst",wind)|grepl("mirco",wind)|grepl("strong",wind)|grepl("gust",wind)]
strong.wind<-strong.wind[!grepl("marine",strong.wind)]
wind<-wind[!(wind %in% high.wind) & !(wind %in% strong.wind)]
wind<-wind[!grepl("tstm",wind) & !grepl("chill",wind) & !grepl("rain",wind) & !grepl("surf",wind) & !grepl("snow",wind) & !grepl("marine",wind) & !grepl("thunder",wind) & !grepl("storm",wind) & !grepl("hail",wind)]
eventtype$EVCAT[which(eventtype$EVTYPE %in% high.wind & eventtype$n==0)]<-"High Wind"
eventtype$EVCAT[which(eventtype$EVTYPE %in% strong.wind & eventtype$n==0)]<-"Strong Wind"
eventtype$EVCAT[which(eventtype$EVTYPE %in% wind & eventtype$n==0)]<-"Wind"

eventtype$n[which(eventtype$EVTYPE %in% high.wind & eventtype$n==0)]<-1
eventtype$n[which(eventtype$EVTYPE %in% strong.wind & eventtype$n==0)]<-1
eventtype$n[which(eventtype$EVTYPE %in% wind & eventtype$n==0)]<-1
rm(wind,high.wind,strong.wind)

#high seas
seas<-unique(eventtype$EVTYPE[grepl("high seas",eventtype$EVTYPE, ignore.case=TRUE) | grepl("heavy seas",eventtype$EVTYPE) | grepl("High Seas",eventtype$EVTYPE) | grepl("rough seas", eventtype$EVTYPE)])
eventtype$EVTYPE[which(eventtype$EVTYPE %in% seas & eventtype$n==0)]<-"High Seas"
eventtype$n[which(eventtype$EVTYPE %in% seas & eventtype$n==0)]<-1
rm(seas)

#Cold, uncat
cold<-unique(eventtype$EVTYPE[grepl("cold",eventtype$EVTYPE,ignore.case = TRUE) | grepl("cool",eventtype$EVTYPE) | grepl("Cold",eventtype$EVTYPE)])
eventtype$EVTYPE[which(eventtype$EVTYPE %in% cold & eventtype$n==0)]<-"Cold"
eventtype$n[which(eventtype$EVTYPE %in% cold & eventtype$n==0)]<-1
rm(cold)

#dry
dry<-unique(eventtype$EVTYPE[grepl("dry",eventtype$EVTYPE, ignore.case = TRUE) | grepl("driest", eventtype$EVTYPE) | grepl("below normal precipitation",eventtype$EVTYPE) | grepl("Dry",eventtype$EVTYPE)])
eventtype$EVTYPE[which(eventtype$EVTYPE %in% dry & eventtype$n==0)]<-"Dry"
eventtype$n[which(eventtype$EVTYPE %in% dry & eventtype$n==0)]<-1
rm(dry)

#uncategorized
eventtype$EVCAT[is.na(eventtype$EVCAT)]<-"Other"
eventtype$n[which(eventtype$n==0)]<-1
```

### Re-Summarize by mortality 
Now that the event types have been categorized, the data are summarized again by the event category. The 10 events with the highest mortality are sub-setted. 
```{r}
eventtype<-eventtype %>% group_by(EVCAT) %>% summarize(Fatalities=sum(FATALITIES),Injuries=sum(INJURIES),health=sum(health,na.rm=TRUE)) %>% arrange(desc(health))

head(eventtype,10)

eventtype.subset<-eventtype[1:10,]
```

# Methods
Fatalities and Injuries were combined into an overall total of health. In this regard, the higher the number, the worse the event was towards population health. The top 10 events most detrimental to public health were isolated and compared.

# Results
In terms of population health, tornados are the detrimental to population health, causing 
5,658 fatalities and 91,364 injuries.  Thunderstorm winds only caused 706 fatalities, and 9,412 injuries. 

### Plot mortalitity rates
The 10 events with the highest mortality are plotted. 
```{r}
g<-ggplot(eventtype.subset, aes(EVCAT,health))
p<-g+geom_col(aes(fill=EVCAT))+labs(title="Mortality Rate of Weather Events in the United States")
print(p)
```
